---
title: "Time-Decay Coefficient Model"
output: pdf_document
---

We only need two packages for this: `tidyverse` for data manipulation and `lme4` for mixed-effects model estimation. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)

team_results <- read_csv("team_results.csv")
```

Creating an exponential time decay function to apply recency weighting to observations in parameter estimation based on `half_life` variable — observations from `half_life` days away from the current date will receive 50% as much weight as present day observations. 

```{r}
time_decay_function <- function(game_date, half_life, current_date = Sys.Date()) {
  
  game_date <- as.Date(game_date)
  current_date <- as.Date(current_date)
  
  days_elapsed <- as.numeric(difftime(current_date, game_date, units = "days"))
  
  decay_value <- (1/2) ^ (days_elapsed / half_life)
  
  return(decay_value)
}
```

The next step is to apply the `half_life` function to create time weights for the model. We filter to only data since 2000 to match the new (reduced) scope of the project.

The chosen half-life initially is 2 years. (This can be tweaked easily in the below code)

```{r}
model_half_life <- 365 * 2

model_data <- team_results %>%
  mutate(time_weight = time_decay_function(date, model_half_life)) %>%
  filter(year(date) >= 2000)
```

Now we fit the models using `lmer`:

```{r}
off_model <- lme4::lmer(score ~ (1 | team) + (1 | opponent) + location, 
                    data = model_data, weights = time_weight)

def_model <- lme4::lmer(opp_score ~ (1 | team) + (1 | opponent) + location, 
                    data = model_data, weights = time_weight)
```

We extract the random effects:

```{r}
off_eff <- ranef(off_model) %>% 
  as.data.frame() %>% 
  filter(grpvar == "team") %>%
  select(team = grp, o_effect = condval, o_sd = condsd)

def_eff <- ranef(def_model) %>% 
  as.data.frame() %>% 
  filter(grpvar == "team") %>%
  select(team = grp, d_effect = condval, d_sd = condsd)

net_eff <- inner_join(off_eff, def_eff, by = "team") %>%
  mutate(net_effect = o_effect - d_effect)
```

Now we have our "team rankings"! (at least a first pass at them)

```{r}
net_eff_clean <- net_eff %>%
  select(team, o_effect, d_effect, net_effect)
```

Top 10 teams in attack, per the model. The `o_effect` term can be interpreted as "how many more goals than the average team do we expect this team to score against their opponent's defense?"

```{r}
net_eff_clean %>%
  arrange(-o_effect) %>%
  head(10)
```


Top 10 teams in defense, per the model. The `d_effect` term can be interpreted as "what's the difference between how many goals this team would allow vs. their opponent and the average team?"

```{r}
net_eff_clean %>%
  arrange(d_effect) %>%
  head(10)
```

Top 10 best teams by "net effect," linear combo of offense and defense:

```{r}
net_eff_clean %>%
  arrange(-net_effect) %>%
  head(10)
```

```{r}
games_played <- team_results %>%
  group_by(team) %>%
  count() %>%
  rename(games_played = n)

net_eff_clean <- net_eff_clean %>%
  left_join(games_played, by = "team")
```

# Match Prediction Visualization

2022 World Cup final: 

```{r Getting Data}
france <- net_eff_clean %>%
  filter(team == "France")

argentina <- net_eff_clean %>%
  filter(team == "Argentina")

avg_goals <- mean(team_results$score)

france_xgoals <- avg_goals + france$o_effect + argentina$d_effect
argentina_xgoals <- avg_goals + argentina$o_effect + france$d_effect
```

```{r Building Match Distribution}
library(bivpois)

get_score_likelihood <- function(team1, team2, go_to = 5) {
  
  goals <- expand.grid(seq(0,go_to), seq(0,go_to))
  
  lik <- map2_vec(.x = goals$Var1, .y = goals$Var2, .f = ~ bivpois::dbp(
    x1 = .x, x2 = .y, lambda = c(team1, team2, 0), logged = FALSE
  ))
  
  goals <- cbind(goals, lik)
  colnames(goals) <- c("ARG", "FRA", "lik")
  return(goals)

}

likelihoods <- get_score_likelihood(argentina_xgoals, france_xgoals)
```

```{r Visualizing Match Distribution}
ggplot(likelihoods, aes(x = ARG, y = FRA)) +
  geom_tile(color = "black", aes(fill = lik), show.legend = FALSE) +
  scale_fill_gradient(low = "white", high = "tomato") +
  geom_text(aes(label = round(lik, 3))) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0,5)) +
  scale_y_continuous(breaks= seq(0,5)) +
  coord_equal() +
  labs(x = "Argentina", y = "France")
```

```{r Visualization Part 2}
likelihoods <- likelihoods %>%
  mutate(result = case_when(
    ARG > FRA ~ "W",
    ARG == FRA ~ "D",
    ARG < FRA ~ "L",
    TRUE ~ NA
  )) %>%
  mutate(lik = ifelse(result == "D", lik * 1.1, lik))

# rescaling to be valid PMF

scale_factor <- 1 / sum(likelihoods$lik)

likelihoods$lik <- likelihoods$lik * scale_factor

ggplot(likelihoods, aes(x = ARG, y = FRA)) +
  geom_tile(color = "black", aes(fill = lik), show.legend = FALSE) +
  scale_fill_gradient(low = "white", high = "tomato") +
  geom_text(aes(label = round(lik, 3))) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0,5)) +
  scale_y_continuous(breaks= seq(0,5)) +
  coord_equal() +
  labs(x = "Argentina", y = "France")
```


```{r Visualization Part 3}

ggplot(likelihoods, aes(x = ARG, y = FRA)) +
  geom_tile(color = "black", aes(fill = result), show.legend = FALSE) +
  scale_fill_manual(values = c("W" = "lightblue", "D" = "grey", "L" = "#0055A4")) +
  geom_text(aes(label = round(lik, 3))) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0,5)) +
  scale_y_continuous(breaks= seq(0,5)) +
  coord_equal() +
  labs(x = "Argentina", y = "France")
```

```{r}
match_probs <- likelihoods %>% 
  group_by(result) %>% 
  summarise(lik_sum = sum(lik)) %>%
  mutate(team = "Argentina") %>%
  mutate(result = factor(result, levels = c("W","D","L"))) %>%
  mutate(team_result = paste(team, result)) %>%
  arrange(result)

library(gt)
library(gtExtras)

gt(match_probs %>% select(team_result, lik_sum)) %>%
  cols_label(team_result = "Result",
             lik_sum = "Odds") %>%
  fmt_percent(lik_sum, decimals = 1) %>%
  gt_theme_538()
```

# Team Ratings

```{r}
net_eff %>% 
  arrange(-net_effect) %>%
  select(team, o_effect, d_effect, net_effect) %>%
  head(10) %>%
  gt() %>%
  cols_label(o_effect = "Offense",
             d_effect = "Defense",
             net_effect = "Total",
             team = "Team") %>%
  fmt_number(o_effect:net_effect, decimals = 2, force_sign = TRUE)
```

# Adjusting for Strength of Opposition

Option 1: across the board

```{r}
net_join <- net_eff %>% 
  select(team, o_effect, d_effect, net_effect)

sos_data <- model_data %>%
  left_join(net_join, by = c("team" = "team")) %>%
  left_join(net_join, by = c("opponent" = "team"), suffix = c("_tm","_opp"))

sos_ratings <- sos_data %>%
  group_by(team) %>%
  summarise(opp_sos = mean(net_effect_opp)) %>%
  arrange(-opp_sos)

adj_coef <- 0.4
sos_net <- left_join(net_eff, sos_ratings, by = "team") %>%
  mutate(adj_net = net_effect + adj_coef * opp_sos,
         adj_eff = adj_net - net_effect) %>%
  mutate(raw_rk = dense_rank(net_effect),
         adj_rk = dense_rank(adj_net)) %>%
  mutate(raw_rk = nrow(net_eff) + 1 - raw_rk,
         adj_rk = nrow(net_eff) + 1 - adj_rk,
         chg_eff = raw_rk - adj_rk) %>%
  arrange(adj_rk)
```

```{r}
sos_data <- model_data %>%
  left_join(net_join, by = c("team" = "team")) %>%
  left_join(net_join, by = c("opponent" = "team"), suffix = c("_tm","_opp"))

model_preds <- sos_data %>%
  mutate(xg = predict(off_model, .),
         xga = predict(def_model, .))
```

```{r}
model_eval <- model_preds %>% 
  select(team, opponent, xg, xga, score, opp_score) %>%
  mutate(xgd = xg - xga, 
         gd = score - opp_score)
```

```{r}
test_eval <- model_eval %>%
  group_by(gd) %>%
  summarise(avg_xgd = mean(xgd), n = n()) %>%
  filter(abs(gd) <= 5)

ggplot(test_eval, aes(x = gd, y = avg_xgd, size = n)) +
  geom_point() +
  coord_equal() +
  scale_x_continuous(limits = c(-5,5), breaks = seq(-5,5)) +
  scale_y_continuous(limits = c(-5,5), breaks = seq(-5,5)) +
  theme_bw() +
  geom_line(size = 1)
```

