---
title: "Time-Decay Coefficient Model"
output: pdf_document
---

We only need two packages for this: `tidyverse` for data manipulation and `lme4` for mixed-effects model estimation. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)

team_results <- read_csv("team_results.csv")
```

Creating an exponential time decay function to apply recency weighting to observations in parameter estimation based on `half_life` variable — observations from `half_life` days away from the current date will receive 50% as much weight as present day observations. 

```{r}
time_decay_function <- function(game_date, half_life, current_date = Sys.Date()) {
  
  game_date <- as.Date(game_date)
  current_date <- as.Date(current_date)
  
  days_elapsed <- as.numeric(difftime(current_date, game_date, units = "days"))
  
  decay_value <- (1/2) ^ (days_elapsed / half_life)
  
  return(decay_value)
}
```

The next step is to apply the `half_life` function to create time weights for the model. We filter to only data since 2000 to match the new (reduced) scope of the project.

The chosen half-life initially is 2 years. (This can be tweaked easily in the below code)

```{r}
model_half_life <- 365 * 2

model_data <- team_results %>%
  mutate(time_weight = time_decay_function(date, model_half_life)) %>%
  filter(year(date) >= 2000)
```

Now we fit the models using `lmer`:

```{r}
off_model <- lme4::lmer(score ~ (1 | team) + (1 | opponent) + location, 
                    data = model_data, weights = time_weight)

def_model <- lme4::lmer(opp_score ~ (1 | team) + (1 | opponent) + location, 
                    data = model_data, weights = time_weight)
```

We extract the random effects:

```{r}
off_eff <- ranef(off_model) %>% 
  as.data.frame() %>% 
  filter(grpvar == "team") %>%
  select(team = grp, o_effect = condval, o_sd = condsd)

def_eff <- ranef(def_model) %>% 
  as.data.frame() %>% 
  filter(grpvar == "team") %>%
  select(team = grp, d_effect = condval, d_sd = condsd)

net_eff <- inner_join(off_eff, def_eff, by = "team") %>%
  mutate(net_effect = o_effect - d_effect)
```

Now we have our "team rankings"! (at least a first pass at them)

```{r}
net_eff_clean <- net_eff %>%
  select(team, o_effect, d_effect, net_effect)
```

Top 10 teams in attack, per the model. The `o_effect` term can be interpreted as "how many more goals than the average team do we expect this team to score against their opponent's defense?"

```{r}
net_eff_clean %>%
  arrange(-o_effect) %>%
  head(10)
```


Top 10 teams in defense, per the model. The `d_effect` term can be interpreted as "what's the difference between how many goals this team would allow vs. their opponent and the average team?"

```{r}
net_eff_clean %>%
  arrange(d_effect) %>%
  head(10)
```

Top 10 best teams by "net effect," linear combo of offense and defense:

```{r}
net_eff_clean %>%
  arrange(-net_effect) %>%
  head(10)
```

